LAK14
========================================================

## Setup

```{r}
# load libraries
library(stringr)
library(ggplot2)
library(topicmodels)
library(tm)
library(slam)
library(lle)
```

## Read Grade 4 data

May consider importing data about the same topic from other classes later.

```{r}
### Read data
notes <- read.csv("KFNotes_HowLightTravels.csv", stringsAsFactors=FALSE)

# remove scaffolds and references
notes$text1 <- gsub("[{].+[}]", " ", notes$text)
notes$text1 <- gsub("\\[\\[.+\\]\\]", " ", notes$text1)

# trying to fix spelling, so LDA training can be more accurate
# there are typos in students' writing too
notes$text1 <- gsub("color", "colour", notes$text1)

# merge title with text
notes$ttext <- paste(notes$title, notes$text1)
```

## Train LDA model

```{r}
## create corpus
corpus <- Corpus(VectorSource(notes$ttext))

# remove some additional stopwords (found them by studying terms under trained topics)
stopwords <- c("can", "does", "will")
corpus <- tm_map(corpus, function(x) 
  removeWords(x, stopwords))

## export corpus to a document-term matrix
dtm <- DocumentTermMatrix(corpus, control = list(
  stemming = TRUE, stopwords = TRUE, minWordLength = 3, 
  removeNumbers = TRUE, removePunctuation = TRUE))

## select the vocabulary based on mean tf-idf
term_tfidf <- tapply(dtm$v/row_sums(dtm)[dtm$i], dtm$j, mean) * 
  log2(nDocs(dtm)/col_sums(dtm > 0))
dtm <- dtm[ , term_tfidf >= 0.1]
dtm <- dtm[row_sums(dtm) > 0, ]

## fit LDA model, k is determined based on Jianwei's paper (15 big threads)
lda <- LDA(dtm, k = 15, control = list(seed = 2013)) # k = number of topics
# save(lda, file="lda.Rdata")
```

## Explore results

### Topics

```{r}
# topic summary
topics <- terms(lda, 5)
words <- apply(topics, 2, function(x) paste(x, collapse = " "))
data.frame(topic=names(words), words=as.vector(words))
```

By studying words under each topic, a lot of topics make sense. Quite nice training results.

For the paper, I will try to read related terms and notes, and try to name each topic. Then I will pick a few topics, and track their progress overtime using the visual below.

### Notes and topics

```{r}
results <- data.frame(id=notes$id, view=notes$view, 
                      topic=factor(topics(lda, 1)), 
                      text=substr(notes$title, 1, 50))
# results

# number of notes under each topic
sort(summary(results$topic), decreasing=TRUE)
```

### Visualization

```{r}
### Locally Linear Embedding (LLE) -- reduce dimension

# Get loading matrix of documents on topics
loading <- posterior(lda, newdata = dtm)

# LLE
X <- loading$topics
calc_k(X, 2, 1, 15, FALSE) # choose k with the smallest rho
results <- lle(X=X, m=2, k=7, reg=2, ss=FALSE, id=TRUE, v=0.9)
time_numeric <- as.numeric(as.POSIXct(strptime(notes$time, "%m/%d/%Y %H:%M:%S")))
time_rank <- rank(time_numeric)
# time_numeric <- as.integer((time_numeric - min(time_numeric)) / 3600 / 24)
points <- data.frame(x=round(results$Y[, 1], 2), 
                     y=round(results$Y[, 2], 2), 
                     id=notes$id,
                     view=notes$view, 
                     author=notes$from_user, 
                     time=time_rank, 
                     topic=factor(topics(lda, 1)),
                     title=notes$title)
# save(points, file="points.Rdata")
```

By topic

```{r}
# qplot(results$id, main="intrinsic dimension")
ggplot(points, aes(x=x,y=y)) + 
  geom_point(data=points, shape=19, aes(x=x, y=y, color=topic), position=position_jitter(w=0.5, h=0.5))
ggsave(file="figure/viz_topic.pdf", width=10, height=8)
```

By view

```{r}
ggplot(points, aes(x=x,y=y)) + 
  geom_point(data=points, shape=19, aes(x=x, y=y, color=view), position=position_jitter(w=0.5, h=0.5))

# add topics
ggplot(points, aes(x=x,y=y)) + 
  geom_point(data=points, aes(x=x, y=y, color=topic, shape=view), size=3, position=position_jitter(w=0.5, h=0.5))
ggsave(file="figure/viz_topic_view.pdf", width=10, height=8)
```

By time

```{r}
ggplot(points, aes(x=x,y=y)) + 
  geom_point(data=points, shape=19, aes(x=x, y=y, color=time), position=position_jitter(w=0.5, h=0.5)) +
  scale_colour_gradient2(mid="red", high="blue")
ggsave(file="figure/viz_time.pdf", width=12, height=8)
```

By students

```{r}
all_author <- 1:nrow(notes)
co_author <- grep(",",notes$from_user)
single_author <- all_author[is.na(match(all_author, co_author))]
points_sub_stu <- points[single_author, ]
points_sub_stu$author <- factor(points_sub_stu$author)

ggplot(points_sub_stu, aes(x=x,y=y)) + 
  geom_point(data=points_sub_stu, shape=19, aes(x=x, y=y, color=author), position=position_jitter(w=0.5, h=0.5))
ggsave(file="figure/viz_stu.pdf", width=12, height=8)
```

### Evolution of topics

```{r}
## create timeseries data
# prepare data needed
notes_ts <- points[, c("time", "topic", "view")]
notes_ts <- notes_ts[with(notes_ts, order(time)), ]

# create time
ts <- data.frame(time=integer(), topic=factor(), count=integer())
topic_nums <- rep(0, length(levels(notes_ts$topic)))
for(i in 1:nrow(notes_ts)) {
  topic <- notes_ts$topic[i]
  topic_nums[topic] <- topic_n <- topic_nums[topic] + 1
  ts <- rbind(ts, c(i, topic, topic_n))
  for(t in 1:length(levels(notes_ts$topic))) {
    if(t != topic) ts <- rbind(ts, c(i, t, topic_nums[t]))
  }
}
names(ts) <- c("time", "topic", "count")
ts$topic <- factor(ts$topic)

# plot results
ggplot(ts, aes(x=time, y=count, colour=topic, group=topic)) + geom_line()
```

Maybe it's better to select a few topics to track, rather than tracking all 15 topics.

```{r}
# here we go -- visualizing a subset of topics, 150/308 of all notes
vt_idx <- c(3, 9, 5, 4, 6)
vt_name <- c("reflection", "refraction", "source of light", "travel", "materials")
top_topics <- data.frame(topic = vt_idx, name = vt_name)
top_topics <- top_topics[with(top_topics, order(topic)), ]

# a subset of note points
notes_ts_sub <- points[points$topic %in% vt_idx, c("time", "topic", "view")]
notes_ts_sub$topic <- factor(notes_ts_sub$topic)

# create time
ts <- data.frame(time=integer(), topic=factor(), count=integer())
topic_nums <- rep(0, length(levels(notes_ts_sub$topic)))
for(i in 1:nrow(notes_ts_sub)) {
  topic <- notes_ts_sub$topic[i]
  topic_nums[topic] <- topic_n <- topic_nums[topic] + 1
  ts <- rbind(ts, c(i, topic, topic_n))
  for(t in 1:length(levels(notes_ts_sub$topic))) {
    if(t != topic) ts <- rbind(ts, c(i, t, topic_nums[t]))
  }
}
names(ts) <- c("time", "topic", "count")
ts$topic <- factor(ts$topic, labels=top_topics$name)

# plot results
ggplot(ts, aes(x=time, y=count, colour=topic, group=topic)) + geom_line()
ggsave(file="figure/viz_topics.pdf", width=12, height=8)
```



Participation in views

```{r}
# create time
ts_v <- data.frame(integer(), integer(), integer())
view_nums <- rep(0, length(levels(notes_ts$view)))
for(i in 1:nrow(notes_ts)) {
  view <- notes_ts$view[i]
  view_nums[view] <- view_n <- view_nums[view] + 1
  ts_v <- rbind(ts_v, c(i, view, view_n))
  for(t in 1:length(levels(notes_ts$view))) {
    if(t != view) ts_v <- rbind(ts_v, c(i, t, view_nums[t]))
  }
}
names(ts_v) <- c("time", "view", "count")
ts_v$view <- factor(ts_v$view, labels=levels(notes_ts$view))

# plot results
ggplot(ts_v, aes(x=time, y=count, colour=view, group=view)) + geom_line()
ggsave(file="figure/viz_views.pdf", width=12, height=8)
```

